---
title: "clustering_prep"
author: "LT"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(factoextra, tidyverse, cluster, stats)
```

# clustering

- standardization is necessary: de-unit is mandatory for clustering so we are comparing apple to apple.

```{r}
# data
data("USArrests")  # Load the data set
df <- USArrests    # Use df as shorter name
df |> head(5)
# remove NA
df <- na.omit(df)
# preprocess
df <- scale(df)
```
## k-means clustering
```{r}
require(stats) # for kmeans()
require(factoextra) # for vizing clustering results

# how to choose the number of clusters (K)?
fviz_nbclust(df, kmeans, method = "wss") +
  geom_vline(xintercept = 4, linetype = 2)

# Compute k-means with k = 4
set.seed(1)
km.res <- kmeans(df, centers = 4, nstart = 25)

# viz kmeans clusters
fviz_cluster(km.res, data = df,
            palette = c("#2E9FDF", "#00AFBB", "#E7B800", "#FC4E07"),
            ellipse.type = "euclid", # Concentration ellipse
            star.plot = F, # Add segments from centroid to items
            repel = TRUE, # Avoid label overplotting (slow)
            ggtheme = theme_minimal()
)
```
- observe for the "elbow"
- Cons
  - requires prior knowledge of the data and prespecified K
  - sensitive to outliers
  - high variances to K
  - ordering rows changes outcomes

## PAM 
- alternative to k-means that is less sensitive to outliers.

```{r}
require(cluster) # for pam()

# determine K
fviz_nbclust(df, pam, method = "silhouette")+
  theme_classic()

# pam
pam.res <- pam(df, 2)

# viz
fviz_cluster(pam.res,
             geom = c("point", "text"),
             palette = c("#00AFBB", "#FC4E07"), # color palette
             ellipse.type = "t", # Concentration ellipse
             repel = TRUE, # Avoid label overplotting (slow)
             ggtheme = theme_classic()
             )
```
## Hierarchical Clustering
- Complete linkage and Ward’s minimal inter-cluster variance method are generally preferred. 
- The higher the height of the fusion, the less similar the objects are.
- remember to verify clustering results by correlating the cophenetic distance to original distance
```{r}
# Compute the dissimilarity matrix
res.dist <- dist(df, method = "euclidean") 
# linkage: take distance matrix and compute dissimilarity among groups
res.hc <- hclust(d = res.dist, method = "ward.D2")

# viz dendrogram
fviz_dend(res.hc, cex = .5)

# Cut in 4 groups and color by groups
fviz_dend(res.hc, k = 4, # Cut in 4 groups
          cex = 0.5, # label size
          k_colors = c("#2E9FDF", "#00AFBB", "#E7B800", "#FC4E07"),
          color_labels_by_k = TRUE, # color labels by groups
          rect = TRUE # Add rectangle around groups
          )
```

```{r}
# horizontal dendrogram
fviz_dend(res.hc, k = 4, 
          cex = 0.4, 
          horiz = TRUE, 
          k_colors = "jco",
          rect = TRUE, 
          rect_border = "jco", 
          rect_fill = TRUE)

# circular/phylogenic
fviz_dend(res.hc, k = 4, 
          cex = 0.4, 
          horiz = TRUE, 
          k_colors = "jco",
          rect = TRUE, 
          rect_border = "jco", 
          rect_fill = TRUE,
          type = "circular")
```
```{r}
# Create a plot of the whole dendrogram,
# and extract the dendrogram data
dend_plot <- fviz_dend(res.hc, k = 4, # Cut in four groups
                       cex = 0.5, # label size
                       k_colors = "jco"
                       )
dend_data <- attr(dend_plot, "dendrogram") # Extract dendrogram data
# Cut the dendrogram at height h = 10
dend_cuts <- cut(dend_data, h = 10)
# Visualize the truncated version containing
# two branches
fviz_dend(dend_cuts$upper)
# plot whole dendrogram
print(dend_plot)
# Plot subtree 1
fviz_dend(dend_cuts$lower[[1]], main = "Subtree 1")
# Plot subtree 2
fviz_dend(dend_cuts$lower[[2]], main = "Subtree 2", 
          type = "circular",  
          repel = TRUE)
```
```{r}
# customization with dendextend
require(dendextend)
# 1. Create a customized dendrogram
mycols <- c("#2E9FDF", "#00AFBB", "#E7B800", "#FC4E07")
dend <- as.dendrogram(res.hc) %>%
    set("branches_lwd", 1) %>% # Branches line width
    set("branches_k_color", mycols, k = 4) %>% # Color branches by groups
    set("labels_colors", mycols, k = 4) %>% # Color labels by groups
    set("labels_cex", 0.5) # Change label size
# 2. Create plot
fviz_dend(dend)
```

```{r}
# save dendrogram
pdf("dendrogram.pdf", width=30, height=15) # Open a PDF
p <- fviz_dend(res.hc, k = 4, cex = 1, k_colors = "jco" ) # Do plotting
print(p)
dev.off()
```

# heatmap
- another way to visualize hierarchical clustering where data values are converted to color scale.

```{r}
require(stats) # heatmap()
require(gplots) # heatmap.2() enhanced heatmap
require(pheatmap) # Draws pretty heatmaps and provides more control to change the appearance
require(d3heamap) # interactive heatmap
require(ComplexHeatmap) # Draws, annotates and arranges complex heatmaps (very useful for genomic)
```

```{r}
df <- scale(mtcars)
# base
heatmap(df, scale = "none")
# Use RColorBrewer color palette names
require("RColorBrewer")
col <- colorRampPalette(brewer.pal(10, "RdYlBu"))(256)
heatmap(df, scale = "none", col = col,
        RowSideColors = rep(c("blue", "pink"), each = 16),
        ColSideColors = c(rep("purple", 5), rep("orange", 6)))
```
```{r}
# enhanced version
require(ComplexHeatmap)
Heatmap(df, 
        name = "mtcars", # title of legend
        column_title = "Variables", 
        row_title = "Samples",
        show_column_names = T,
        show_row_names = T,
        show_column_dend = T,
        show_row_dend = T,
         clustering_distance_rows = "euclidean",
        clustering_method_rows = "ward.D2",
        row_names_gp = gpar(fontsize = 7)) # Text size for row names)

# Splitting heatmap by rows
Heatmap(df, 
        split = mtcars$cyl, 
        name = "mtcars", # title of legend
        row_names_gp = gpar(fontsize = 7))

# Split by combining multiple variables
Heatmap(df, name ="mtcars",
        split = data.frame(cyl = mtcars$cyl, am = mtcars$am))

############################################################
# annotation
############################################################
# Annotation data frame
annot_df <- data.frame(cyl = mtcars$cyl, 
                       am = mtcars$am,
                       mpg = mtcars$mpg)
# Define colors for each levels of qualitative variables
# Define gradient color for continuous variable (mpg)
col = list(cyl = c("4" = "green", "6" = "gray", "8" = "darkred"),
           am = c("0" = "yellow", "1" = "orange"),
           mpg = circlize::colorRamp2(c(17, 25),
                                      c("lightblue", "purple")) 
           )
# Create the heatmap annotation
ha <- HeatmapAnnotation(df = annot_df, col = col, 
                        show_legend = F)
# Combine the heatmap and the annotation
Heatmap(t(df), name = "mtcars",
        top_annotation = ha)
```

```{r}
# complex Annotation 
df <- t(df)
# Define some graphics to display the distribution of columns
.hist = anno_histogram(df, 
                       gp = gpar(fill = "lightblue"))
.density = anno_density(df, type = "line", 
                        gp = gpar(col = "blue"))

ha_mix_top = HeatmapAnnotation(hist = .hist, 
                               density = .density, 
                               which = "column",
                               height = unit(3, "cm"))
# Define some graphics to display the distribution of rows
.violin = anno_density(df, type = "violin",
                       gp = gpar(fill = "lightblue"), 
                       which = "row")
.boxplot = anno_boxplot(df, which = "row")

ha_mix_right = HeatmapAnnotation(violin = .violin, 
                                 bxplt = .boxplot,
                                 which = "row", 
                                 width = unit(4, "cm"))
# Combine annotation with heatmap
Heatmap(df, name = "mtcars",
        column_names_gp = gpar(fontsize = 8),
        top_annotation = ha_mix_top) + 
  ha_mix_right
```

































