---
title: "proj3_classification_biclass"
author: "LT"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(caret, mlbench, tidyverse)
```

# DEFINE PROBLEM

we will investigate the Wisconsin Breast Cancer Data here, and classify tumor status using the design matrix.

  - LOAD PACKAGES
  - LOAD DATASET
  - DATA SPLIT
  
```{r}
# load data
data("BreastCancer")
# split data
# create a list of 80% of the rows in the original dataset we can use for training
set.seed(7)
validationIndex <- createDataPartition(BreastCancer$Class, p=0.80, list=FALSE)
# select 20% of the data for validation
validation <- BreastCancer[-validationIndex,]
# use the remaining 80% of data to training and testing the models
dataset <- BreastCancer[validationIndex,]
```

# SUMMARIZE DATA
  - DESCRIPTIVE STATISTICS
  - DATA VISUALIZATIONS
# PREPARE DATA
  - DATA CLEANING
  - FEATURE SELECTION
  - DATA ENGINEERING/TRANSFORMATION
# EVALUATE ALGORITHMS
  - TEST OPTIONS AND EVALUATION METRIC
  - SPOT-CHECK ALGORITHMS
  - COMPARE ALGORITHMS
# IMPROVE RESULTS
  - ALGORITHM TUNING
  - ENSEMBLES
# PRESENT RESULTS
  - PREDICTIONS ON VALIDATION SET
  - CREATE STANDALONE MODEL ON ENTIRE TRAINING SET
  - SAVE MODEL FOR LATER USE
















































