---
title: "classification_multiclass"
author: "LT"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(caret, mlbench)
```

## DEFINE PROBLEM & LOAD DATA
```{r}
# attach the iris dataset to the environment
data(iris)
# rename the dataset
dataset <- iris
# create a list of 80% of the rows in the original dataset we can use for training
validationIndex <- createDataPartition(dataset$Species, p=0.80, list=FALSE)
# select 20% of the data for validation
validation <- dataset[-validationIndex,]
# use the remaining 80% of data to training and testing the models
dataset <- dataset[validationIndex,]
```

## SUMMARIZE DATA
- peak at data
```{r}
# dimensions of dataset
dim(dataset)
# list types for each attribute
sapply(dataset, class)
# list the levels for the class
levels(dataset$Species)
# summarize the class distribution
percentage <- prop.table(table(dataset$Species)) * 100
cbind(freq=table(dataset$Species), percentage=percentage)
```
- statistical summary
```{r}
# summarize attribute distributions
summary(dataset)
# skewness
apply(dataset[,1:4], 2, skewness)
```
- visualization
```{r}
# univariate
# split input and output
x <- dataset[,1:4]
y <- dataset[,5]
# boxplot for each attribute on one image
par(mfrow=c(1,4))
for(i in 1:4) {
  boxplot(x[,i], main=names(iris)[i])
}

# barplot for class breakdown
plot(y)


# multivariate plot
caret::featurePlot(x=x, y=y, plot="ellipse")
caret::featurePlot(x=x, y=y, plot="box")

# density plots for each attribute by class value
scales <- list(x=list(relation="free"), y=list(relation="free"))
featurePlot(x=x, y=y, plot="density", scales=scales)
```

## PREPARE DATA

## EVALUATE ALGORITHMS
time to create some models of the data and estimate their accuracy on unseen data. 

- apply 10-fold CV without repeat
- build 5 models to classify flower species
  - LDA
  - Classification and Regression Trees (CART)
  - SVM with a radial kernel
  - RF
- select best model

```{r model_build}
# Run algorithms using 10-fold cross validation
trainControl <- trainControl(method="cv", number=10)
metric <- "Accuracy"

# LDA
set.seed(1)
fit.lad <- train(Species ~ ., data = dataset, 
                 method ="lda", 
                 metric = metric,
                 trControl = trainControl)

# CART
set.seed(1)
fit.cart <- train(Species ~ ., data = dataset, 
                 method ="rpart", 
                 metric = metric,
                 trControl = trainControl)

# KNN
set.seed(1)
fit.knn <- train(Species ~ ., data = dataset, 
                 method ="knn", 
                 metric = metric,
                 trControl = trainControl)

# SVM
set.seed(1)
fit.svm <- train(Species ~ ., data = dataset, 
                 method ="svmRadial", 
                 metric = metric,
                 trControl = trainControl)

# RF
set.seed(1)
fit.rf <- train(Species ~ ., data = dataset, 
                 method ="rf", 
                 metric = metric,
                 trControl = trainControl)
```

```{r model_comparison}
# summarize accuracy of models
results <- resamples(list(lda=fit.lad, cart=fit.cart, 
                          knn=fit.knn, svm=fit.svm, rf=fit.rf))
summary(results)
# plot accuracy of models
dotplot(results)
```

```{r best_model}
# summarize Best Model
print(fit.lad)
```

## IMPROVE RESULT


## PREDICTIONS
```{r}
# estimate skill of LDA on the validation dataset
predictions <- predict(fit.lad, validation)
confusionMatrix(predictions, validation$Species)
```



































